---
title: "FD_ordination"
author: "Aimara Planillo"
date: "2024-01-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functional diversity and PCoA


```{r prepare workspace}
source("./rscripts/source_packages.R")
```

### load data
```{r}
## grid data
bird_data <- read.csv("./data/grid_birds_abundance_corrected.csv") %>%
  rename(transect = X) %>% 
  mutate(transect = tolower(transect)) %>% 
  column_to_rownames("transect")
 
ncol(bird_data) #95 species
head(bird_data)

bird_matrix <- as.matrix(bird_data)

## order species alphabetically
bird_matrix2 <- bird_matrix %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("species") %>% 
  arrange(species) %>% 
  column_to_rownames("species") %>% 
  t()


## trait data
birds_traits <- read.csv("data/birds_all_traits.csv") 
head(birds_traits)
birds_traits <- birds_traits %>% 
  arrange(species) %>% 
  filter(!species %in% c("Apus apus", "Passer domesticus")) %>% 
  mutate(migrant_num = as.numeric(migrant_num),
         diet_cat = as.factor(diet_cat))%>%
  dplyr::select(-migrant_cat)

# get species names with underscore
name1 <- stringr::word(birds_traits$species, 1)
name2 <- stringr::word(birds_traits$species, 2)

new_names <- paste(name1,name2, sep = "_")

traits_matrix <- birds_traits %>%
  dplyr::select(-species, -sp_synonym) 
rownames(traits_matrix) <- new_names
```

# 1. Calculate FD indices

```{r}
FD_birds <- FD::dbFD(x = traits_matrix, a = bird_matrix2, 
                     w.abun = TRUE, corr = 'cailliez', 
                     m = 'max', clust.type = 'ward',
                     print.pco = TRUE, calc.FGR = FALSE, ## since we decided we'll focus on FRic, Fdis 
                     stand.FRic = TRUE, calc.FDiv = TRUE)  

FD_birds
# saveRDS(FD_birds, "./output/fd_birds.rds")
# FD_birds <- readRDS("./output/fd_birds.rds")

## weighted mean of functional traits
FD_birds$CWM


mean_traits_transect <- FD_birds$CWM
summary(mean_traits_transect)
table(mean_traits_transect$diet_cat)

# extract diversity values
# Add info into one table
birds_fr <- FD_birds$FRic 
birds_fe <-FD_birds$FEve
birds_fd <- FD_birds$FDiv

birds_fr <-as.data.frame(birds_fr)
birds_fe <-as.data.frame(birds_fe)
birds_fd <-as.data.frame(birds_fd)

row.names(birds_fr)
birds_fr$transect <- rownames(birds_fr)
birds_fe$transect <- rownames(birds_fe)
birds_fd$transect <- rownames(birds_fd)

birds_functional_measures <- left_join(birds_fr, birds_fe,by = "transect")
birds_functional_measures <- left_join(birds_functional_measures, birds_fd, by = "transect")
head(birds_functional_measures)
birds_functional_measures <- birds_functional_measures %>% 
  dplyr::select(transect, everything())

# write.csv(birds_functional_measures, "./output/functional_measures_transects.csv", row.names = FALSE)
```


### summaries
```{r}
birds_functional_measures %>% 
  summarise(across(where(is.numeric), list(mean = mean, sd = sd, min = min, max = max)))
```
  birds_fr_mean birds_fr_sd birds_fr_min birds_fr_max birds_fe_mean birds_fe_sd birds_fe_min birds_fe_max birds_fd_mean birds_fd_sd birds_fd_min birds_fd_max
1   0.009612069  0.02091402 2.136223e-09    0.1043922       0.66899   0.0528327    0.5615259    0.7639633     0.8893223  0.02152722     0.852556     0.948475


# 2. PCoA

```{r load data}
## bird abund transect
bird_abund <- bird_data %>% 
  rownames_to_column("transect")

## trait data transect
mean_traits_transect <- FD_birds$CWM %>% 
  rownames_to_column("transect")

## env covariates grid
covariates_grid <- read.csv("./data/grids_environmental_variables.csv") 

## env covariates territory
covariates_terr <- read.csv("./data/territories_environmental_variables.csv") 
```



## 2.1 PCoA grid scale 

### prepare data for grid scale 
```{r}
## merge together to make sure the order is correct
all_data_grid <- bird_abund %>% 
  left_join(covariates_grid, by = "transect") %>% 
  left_join(mean_traits_transect, by = "transect") %>% 
  column_to_rownames("transect")

nrow(all_data_grid)
# [1] 29 sites
ncol(all_data_grid)
# [1] 112 # 95 species abundances, 2 coordinates, 10 env variables and 5 and mean trait values

## divide into the proper tables
species_matrix <- all_data_grid[,1:95]
env_matrix <- all_data_grid[,98:107]
traits_matrix <-all_data_grid[,108:112]

str(species_matrix)
str(env_matrix)
str(traits_matrix)
```

### Run ordination
```{r}
## species ordination
PCoA_communitycomposition <- capscale(species_matrix ~ 1, dist="bray", metaMDS = TRUE)

## add environmetal variables
ef <- envfit(PCoA_communitycomposition, env_matrix, permu = 999, na.rm = T)
ef

## Add traits 
tq <- envfit(PCoA_communitycomposition, traits_matrix, permu = 999)

# Add species 
spf <- envfit(PCoA_communitycomposition, species_matrix, permu = 999)
spf

plot(PCoA_communitycomposition, type = "t", display = "sites")
plot(ef, cex = 0.7)
plot(tq, cex = 0.6, col = "darkgreen")
plot(spf, col = "red")
```

```{r extract data for plotting}
spf$vectors$arrows
spf$vectors$pvals

sign_sp <- names(spf$vectors$pvals[spf$vectors$pvals < 0.05]) # get names of the significant sp

species_scores <- spf$vectors$arrows %>%
  as.data.frame() %>%
  rownames_to_column("species") 
species_scores <- filter(species_scores, species %in% sign_sp)

# get sites loads in axis
site_scores <- as.data.frame(scores(PCoA_communitycomposition)$sites)

# get coordinates of environmental variables
env_coord_cont <- as.data.frame(scores(ef, "vectors")) * ordiArrowMul(ef)
env_variables <- c("imperv", "pop_den", "tcd", "light", "noise", "veg_pca1", "veg_pca2", "ndvi", "glcm", "glcm_hmg")
rownames(env_coord_cont) <- env_variables

# remove ndvi because overlaps almost perfectly with glcm
env_coord_cont <- env_coord_cont[-8,] 

# get coordinates of traits 
traits_coord_cat <- as.data.frame(scores(tq, "factors"))
rownames(traits_coord_cat) <- c("Invertebrates", "Omnivore")
traits_coord_cat

traits_coord_cont <- as.data.frame(scores(tq, "vectors")) * ordiArrowMul(tq)
rownames(traits_coord_cont) <- c("Migrant", "BodyMass", "Longevity", "Clutch_size")
traits_coord_cont
```

```{r ggplot}
## plot in ggplot
(myplot <- ggplot(data = site_scores, aes(x = MDS1, y = MDS2)) + 
  geom_point(colour = "orange", size = 5, alpha = 0.3) + 
  # scale_colour_manual(values = c( "orange")) + 
  geom_segment(data = env_coord_cont, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "darkgreen", 
               arrow = arrow(length = unit(0.3,"cm"))) +
  geom_text(data = env_coord_cont, aes(x = MDS1*1.2, y = MDS2*1.2),size = 4, colour = "darkgreen", 
            fontface = "bold", label = row.names(env_coord_cont)) + 
  geom_segment(data = traits_coord_cat, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "black", 
               arrow = arrow(length = unit(0.3,"cm"))) +
  geom_text(data = traits_coord_cat, aes(x = MDS1*1.2, y = MDS2*1.2),size = 4,  colour = "black", 
            fontface = "bold", label = row.names(traits_coord_cat)) + 
  geom_segment(data = traits_coord_cont, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "black", 
               arrow = arrow(length = unit(0.3,"cm"))) +
  geom_text(data = traits_coord_cont, aes(x = MDS1*1.2, y = MDS2*1.2),size = 4,  colour = "black", 
            fontface = "bold", label = row.names(traits_coord_cont)) + 
   geom_hline(yintercept = 0, col = "grey70", lty = 4) +
  geom_vline(xintercept = 0, col = "grey70", lty = 4) +

  # geom_segment(data = species_scores, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
  #              size =0.1, alpha = 0.5, colour = "brown1") +
  # geom_text(data = species_scores, aes(x = MDS1, y = MDS2), size = 2.5, colour = "brown1", 
  #           alpha = 0.6,fontface = "italic", label = species_scores$species) + 
  xlim(c(-1.5,1.5)) +
  ylim(c(-1.5,1.5)) +
  theme(axis.title = element_text(size = 14, face = "bold", colour = "black"), 
       panel.background = element_blank(), 
       panel.border = element_rect(fill = NA, colour = "black"), 
        # axis.ticks = element_blank(), axis.text = element_blank(), legend.key = element_blank(), 
        # legend.title = element_text(size = 10, face = "bold", colour = "grey30"), 
        # legend.text = element_text(size = 9, colour = "grey30")), 
       axis.text = element_text(size = 14, colour = "black")))

ggsave(plot = myplot, 
       filename = "./plots/PCoA_grid_scale.png", 
       dpi = 900, height = 8, width = 8)
```

# 2.2 PCoA territory scale


### prepare data for territory scale 
```{r}
covariates_terr_mean <- covariates_terr %>% 
  dplyr::select(-nvisits) %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE))))

## merge together to make sure the order is correct


all_data_terr <- bird_abund %>% 
  left_join(covariates_terr_mean, by = "transect") %>% 
  left_join(mean_traits_transect, by = "transect") %>% 
  column_to_rownames("transect")

nrow(all_data_terr)
# [1] 29 sites
ncol(all_data_terr)
# [1] 110 # 95 species abundances, 10 env variables and 5 and mean trait values

## divide into the proper tables
species_matrix_terr <- all_data_terr[,1:95]
env_matrix_terr <- all_data_terr[,96:105]
traits_matrix_terr <-all_data_terr[,106:110]

str(species_matrix_terr)
str(env_matrix_terr)
str(traits_matrix_terr)
```


```{r PCoA territories - bird community composition}
# Run PCoa of sites based on community compositon (species abundances)
PCoA_terr <- capscale(species_matrix_terr ~ 1, dist="bray", metaMDS = TRUE)

# Add environmental variables
ef_terr <- envfit(PCoA_terr, env_matrix_terr, permu = 999, na.rm = T)

# Add traits per quadrant
tq_terr <- envfit(PCoA_terr, traits_matrix_terr, permu = 999)

# Add species
spf_terr <- envfit(PCoA_terr, species_matrix_terr, permu = 999)

# # Get significant species loads in the axes
spf_terr$vectors$arrows
spf_terr$vectors$pvals

sign_sp_terr <- names(spf_terr$vectors$pvals[spf_terr$vectors$pvals < 0.05]) # get names of the significant sp

species_scores_terr <- spf_terr$vectors$arrows %>%
  as.data.frame() %>%
  rownames_to_column("species") 
species_scores_terr <- filter(species_scores_terr, species %in% sign_sp_terr)

# get sites loads in axis
site_scores_terr <- as.data.frame(scores(PCoA_terr)$sites)

# get coordinates of environmental variables
env_coord_cont_terr <- as.data.frame(scores(ef_terr, "vectors")) * ordiArrowMul(ef_terr)

env_variables <- c("imperv", "pop_den", "tcd", "light", "noise", "veg_pca1", "veg_pca2", "ndvi", "glcm", "glcm_hmg")
rownames(env_coord_cont_terr) <- env_variables

# remove ndvi because overlaps almost perfectly with glcm
env_coord_cont_terr <- env_coord_cont_terr[-8,] 

# get coordinates of traits 
traits_coord_cat_terr <- as.data.frame(scores(tq_terr, "factors"))
rownames(traits_coord_cat_terr) <- c("Invertebrates", "Omnivore")
traits_coord_cat_terr

traits_coord_cont_terr <- as.data.frame(scores(tq_terr, "vectors")) * ordiArrowMul(tq_terr)
rownames(traits_coord_cont_terr) <- c("Migrant", "BodyMass", "Longevity", "Clutch_size")
traits_coord_cont_terr
```

```{r plot PCoA - territories}
(myplot2 <- ggplot(data = site_scores_terr, aes(x = MDS1, y = MDS2)) + 
  geom_point(colour = "orange", size = 5, alpha = 0.3) + 
  # scale_colour_manual(values = c( "orange")) + 
  geom_segment(data = env_coord_cont_terr, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "darkgreen", 
               arrow = arrow(length = unit(0.3,"cm"))) +
  geom_text(data = env_coord_cont_terr, aes(x = MDS1*1.2, y = MDS2*1.2),size = 4, colour = "darkgreen", 
            fontface = "bold", label = row.names(env_coord_cont_terr)) + 
  geom_segment(data = traits_coord_cat_terr, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "black", 
               arrow = arrow(length = unit(0.3,"cm"))) +
  geom_text(data = traits_coord_cat_terr, aes(x = MDS1*1.2, y = MDS2*1.2),size = 4,  colour = "black", 
            fontface = "bold", label = row.names(traits_coord_cat_terr)) + 
  geom_segment(data = traits_coord_cont_terr, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "black", 
               arrow = arrow(length = unit(0.3,"cm"))) +
  geom_text(data = traits_coord_cont_terr, aes(x = MDS1*1.2, y = MDS2*1.2),size = 4,  colour = "black", 
            fontface = "bold", label = row.names(traits_coord_cont_terr)) + 
   geom_hline(yintercept = 0, col = "grey70", lty = 4) +
  geom_vline(xintercept = 0, col = "grey70", lty = 4) +

  # geom_segment(data = species_scores_terr, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
  #              size =0.1, alpha = 0.5, colour = "brown1") +
  # geom_text(data = species_scores_terr, aes(x = MDS1, y = MDS2), size = 2.5, colour = "brown1", 
  #           alpha = 0.6,fontface = "italic", label = species_scores_terr$species) + 
  xlim(c(-1.5,1.5)) +
  ylim(c(-1.5,1.5)) +
  theme(axis.title = element_text(size = 14, face = "bold", colour = "black"), 
       panel.background = element_blank(), 
       panel.border = element_rect(fill = NA, colour = "black"), 
        # axis.ticks = element_blank(), axis.text = element_blank(), legend.key = element_blank(), 
        # legend.title = element_text(size = 10, face = "bold", colour = "grey30"), 
        # legend.text = element_text(size = 9, colour = "grey30")), 
       axis.text = element_text(size = 14, colour = "black")))

ggsave(plot = myplot2, 
       filename = "./plots/PCoA_territory_scale.png", 
       dpi = 900, height = 8, width = 8)
```





##########################
## PCoA territory scale ##  
##########################

##############################
##PCoA with mean values of environmental variables for territories
terr_mean_values <- read.csv(paste0(workdir, "/data_raw/territories_mean_env_values.csv"), header = T)

env_data_terr_matrix <-terr_mean_values %>%
  column_to_rownames("routcode")

###PCoA
### PCoa of sites based on community compositon (species abundances)
PCoA_territorycomposition <- capscale(species_matrix ~ 1, dist="bray", metaMDS = TRUE)
plot(PCoA_territorycomposition)

# Add environmental variables
ef_terr <- envfit(PCoA_territorycomposition, env_data_terr_matrix, permu = 999, na.rm = TRUE)
ef_terr

plot(PCoA_territorycomposition, type = "t", display = "sites", cex = 1)
plot(ef_terr, cex = 0.7)

#add traits
plot(PCoA_communitycomposition, type = "t", display = "sites", cex = 1)
plot(tq, cex = 0.6, col = "darkgreen")

# Add species (although not needed)
spf_terr <- envfit(PCoA_territorycomposition, species_matrix, permu = 999)
spf_terr

plot(PCoA_territorycomposition, type = "t", display = "sites")
plot(spf_terr, col = "red", cex = 0.7)

###with traits in ggplot
site_scores_terr <- as.data.frame(scores(PCoA_territorycomposition)$sites)
site_scores_terr$colors <- c(rep("sites", 29)) # This is just and example how to color the sites

sign_sp <- names(spf$vectors$pvals[spf$vectors$pvals < 0.05]) # get names of the significant sp

species.scores <- spf$vectors$arrows %>%
  as.data.frame() %>%
  rownames_to_column("species") 
species.scores <- filter(species.scores, species %in% sign_sp)

env_coord_cont_terr <- as.data.frame(scores(ef_terr, "vectors")) * ordiArrowMul(ef_terr)

traits_coord_cont <- as.data.frame(scores(tq, "factors"))

log_body_mass <- as.data.frame(scores(tq, "vectors")) * ordiArrowMul(tq)

myplot3 <- ggplot(data = site_scores_terr, aes(x = MDS1, y = MDS2)) + 
  geom_point(aes(colour = colors), size = 3, alpha = 0.5) + 
  scale_colour_manual(values = c( "darkblue")) + 
  geom_segment(data = env_coord_cont_terr, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "grey30") +
  geom_text(data = env_coord_cont_terr, aes(x = MDS1, y = MDS2),size = 3,  colour = "grey30", 
            fontface = "bold", label = row.names(env_coord_cont_terr)) + 
  geom_segment(data = traits_coord_cont, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "forestgreen") +
  geom_text(data = traits_coord_cont, aes(x = MDS1, y = MDS2),size = 3,  colour = "forestgreen", 
            fontface = "bold", label = row.names(traits_coord_cont)) + 
  geom_segment(data = log_body_mass, aes(x = 0, y = 0, xend = MDS1*0.5, yend = MDS2*0.5), 
               size =1, alpha = 0.5, colour = "darkgreen") +
  geom_text(data = log_body_mass, aes(x = MDS1*0.5, y = MDS2*0.5),size = 3,  colour = "darkgreen", 
            fontface = "bold", label = row.names(log_body_mass)) + 
  geom_segment(data = species.scores, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "brown1") +
  geom_text(data = species.scores, aes(x = MDS1, y = MDS2), size =2.5, colour = "brown1", 
            fontface = "italic",alpha = 0.6, label = species.scores$species) + 
  xlim(c(-3,2.5)) +
  ylim(c(-1.5,1.5)) +
  theme(axis.title = element_text(size = 10, face = "bold", colour = "grey30"), 
        panel.background = element_blank(), panel.border = element_rect(fill = NA, colour = "grey30"), 
        axis.ticks = element_blank(), axis.text = element_blank(), legend.key = element_blank(), 
        legend.title = element_text(size = 10, face = "bold", colour = "grey30"), 
        legend.text = element_text(size = 9, colour = "grey30")) +
  labs(colour = "")
myplot3

myplot4 <- ggplot(data = site_scores_terr, aes(x = MDS1, y = MDS2)) + 
  geom_point(aes(colour = colors), size = 3, alpha = 0.5) + 
  scale_colour_manual(values = c( "darkblue")) + 
  geom_segment(data = env_coord_cont_terr, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "grey30") +
  geom_text(data = env_coord_cont_terr, aes(x = MDS1, y = MDS2),size = 3,  colour = "grey30", 
            fontface = "bold", label = row.names(env_coord_cont_terr)) + 
  geom_segment(data = traits_coord_cont, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "forestgreen") +
  geom_text(data = traits_coord_cont, aes(x = MDS1, y = MDS2),size = 3,  colour = "forestgreen", 
            fontface = "bold", label = row.names(traits_coord_cont)) + 
  geom_segment(data = log_body_mass, aes(x = 0, y = 0, xend = MDS1, yend = MDS2), 
               size =1, alpha = 0.5, colour = "darkgreen") +
  geom_text(data = log_body_mass, aes(x = MDS1, y = MDS2),size = 3,  colour = "darkgreen", 
            fontface = "bold", label = row.names(log_body_mass)) + 
  theme(axis.title = element_text(size = 10, face = "bold", colour = "grey30"), 
        panel.background = element_blank(), panel.border = element_rect(fill = NA, colour = "grey30"), 
        axis.ticks = element_blank(), axis.text = element_blank(), legend.key = element_blank(), 
        legend.title = element_text(size = 10, face = "bold", colour = "grey30"), 
        legend.text = element_text(size = 9, colour = "grey30")) +
  labs(colour = "Quadrants")
myplot4



