---
title: "FilteredData"
author: "Aimara Planillo"
date: "2024-01-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#######
# NEW #
#######






#################################
#################################
#################################
check the code below

# Clean and Prepare data for bird analysis

This script loads and cleans the bird data for JSDM
Remove all the species that are
- rare (less than 3 sites)
- aquatic
- exotic

```{r packages}
library(dplyr) # Working with data
library(sf)
```


## Workspace
```{r}
WorkDir <- getwd()

Data_wd <- file.path(WorkDir, "2_Raw_data/")
Output_wd <- file.path(WorkDir, "4_Processed_data/")
```


# Correct bird data by TRANSECT LENGTH

To correct the number of observations in each transect, we divide the number of individuals by the length of the transect

```{r load data}
transects <- read_sf(paste0(Data_wd,
                            "/geodata//birds_transects_geometry_25833.gpkg"))
transects
plot(st_geometry(transects))
# extract length
transects.lengths <- st_length(transects)
transects.birds <- cbind.data.frame(site = transects$routcode_2, 
                                    length = as.numeric(transects.lengths))

# Bird data
birds.raw <- read.csv(paste0(Data_wd, "/Birds_sitexsp_fullnames.csv"))
head(birds.raw)

# Traid data
birds.traits <- read.csv(paste0(Data_wd,  "/Birds_traits.csv"))
colnames(birds.traits) <- tolower(colnames(birds.traits))
head(birds.traits)

# Create a new column with the scientific names in the same format as the dataframe with the abundance
birds.traits$scientific2 <- as.character(birds.traits$scientific)

scientific.names <- as.character(birds.traits$scientific)
m <- list()
for (i in 1:length(scientific.names)){
  m <- strsplit(scientific.names, " ")[i]
  birds.traits$scientific2[i] <- paste0(m[[1]][1], "_", m[[1]][2])
}

head(birds.traits)

# Set birds raw data to the same order as they appear in the traits (alphabetical)
order.alph <- as.character(c("SITE", birds.traits$scientific2))

birds.raw <- birds.raw[order.alph]
head(birds.raw)

# Reorder the transects in the same order as the bird data and name them in the same way

birds.raw$SITE
transects.birds$site <- as.factor(toupper(transects.birds$site))

transects.birds <- transects.birds[order(transects.birds$site),]
transects.birds
```


```{r Correct bird data by transect length}
birds.raw.corrected <- birds.raw
for (i in 2:ncol(birds.raw)){
  birds.raw.corrected[,i] <- birds.raw[,i] / transects.birds$length *1000 #we obtain number per km
}
birds.raw.corrected

# double check numbers are ok:
turdus <- as.numeric(birds.raw$Turdus_merula)
lengths <- as.numeric(transects.birds$length)
turdus/lengths * 1000
```

```{r remove introduced species}
introduced.sp <- birds.traits$scientific2[birds.traits$introduced == "Introduced"]
introduced.sp

birds.native <- select(birds.raw, -introduced.sp)
head(birds.native)
nrow(birds.native)
ncol(birds.native) -1 # -1 because of site column
```

```{r remove rare species}
# Remove species that were located only in one (1) or one and two sites (2)
colSums(birds.native[,2:ncol(birds.native)] != 0) >2

birds.norare2 <- birds.native[2:ncol(birds.native)] [colSums(birds.native[,2:ncol(birds.native)] != 0) > 2]

ncol(birds.native) -1
ncol(birds.norare2)
colSums(birds.native[,2:ncol(birds.native)] != 0) >2

# get the species remaining
birds.sp.norare2 <- colnames(birds.norare2[,1:ncol(birds.norare2)])
length(birds.sp.norare2)

# get the traits only for those
birds.traits.norare2 <- birds.traits[birds.traits$scientific2 %in% birds.sp.norare2,]
nrow((birds.traits.norare2)) # [1] 70
```

```{r remove aquatic species}
# Remove species related to water
aquatic2.sp <- birds.traits.norare2$scientific2[birds.traits.norare2$foraging_habitat == "freshwater"]
aquatic2.sp

birds.norare2.noaquatic <- select(birds.norare2, -aquatic2.sp)
birds.norare2.noaq.sp <- colnames(birds.norare2.noaquatic[1:ncol(birds.norare2.noaquatic)])
head(birds.norare2.noaquatic)
ncol(birds.norare2.noaquatic) # [1] 66
```


# Save final clean tables
```{r}
# Birds Without exotic, species in 1 or 2 sites and aquatic sp corrected by transect length
birds.norare2.noaq.corrected <- birds.raw.corrected %>%
  select(., colnames(birds.norare2.noaquatic[1:ncol(birds.norare2.noaquatic)-1])) %>%
  cbind(., site = birds.raw.corrected$SITE)

ncol(birds.norare2.noaq.corrected)
head(birds.norare2.noaq.corrected)

write.csv (birds.norare2.noaq.corrected, 
           paste0(Output_wd, "/NEWBirds_abund_noexo_norare2_noaquatic_bykm.csv"), row.names = FALSE)



# Traits for filtered species
traits.norare2.noaquatic <- birds.traits[birds.traits$scientific2 %in% birds.norare2.noaq.sp,]
nrow(traits.norare2.noaquatic)

write.csv(traits.norare2.noaquatic, paste0(Output_wd, "/NEWBirds_traits_noexo_norare2_noaquatic.csv"), row.names = FALSE)
```


# Summaries of the bird data

* Three Traits: 
  ~ "migrant" + "diet.inv" + "log10.bodymass" 
  
* Birds species filtered:
- No exotic sp
- No species in only 1 or 2 sites
- No aquatic species
- Species abundance corrected by transect length


```{r summary of bird abundances and prevalence}
# load bird data
birds_km <- read.csv("./4_processed_data/NEWBirds_abund_noexo_norare2_noaquatic_bykm.csv") # matrix with sites in rows and bird abund in columns
birds_km$site <- tolower(birds_km$site)
head(birds_km)
ncol(birds_km)

## Average species numbers
bird_data <- birds_km %>%
  select(-site) 
  
bird_species <- rowSums(bird_data > 0)
bird_species

summary(bird_species)

## Prevalence - number of sites for each species
colSums(bird_data > 0)

prevalence <- data.frame(species = colnames(bird_data),  n.sites = colSums(bird_data > 0))
prevalence$percent <- prevalence$n.sites/29 * 100
prevalence

colSums(prevalence[3] > 79.9) # Number of species in 80% or more sites
colSums(prevalence[3] == 100) # Number of speciesin all sites


## Abundaces

# abundance in the sites present
bird.avgabund <- colSums(bird_data) / prevalence$n.sites

birds.summary <- cbind.data.frame(prevalence, avgabund = bird.avgabund)
plot(birds.summary$avgabund~birds.summary$n.sites)
birds.summary
 
summary(birds.summary$avgabund)
birds.summary[birds.summary$avgabund > 15,] # species with mean abundance > 15
birds.summary[birds.summary$avgabund == min(birds.summary$avgabund),] # species with min mean abund


### Check distribution of species abundance
# histograms of the total abundance of the species
hist(colSums(bird_data))
# histograms of the sp with higher abundance in each taxa
hist(bird_data[,which.max(colSums(bird_data))])


```


## Summary traits for the selected species to do the analysis

```{r summmary table traits}
# Load traits
Trdata.tmp <- read.csv("./4_Processed_data/NEWBirds_traits_noexo_norare2_noaquatic.csv") # matrix with species in rows and traits in columns
Trdata.tmp
nrow(Trdata.tmp)
ncol(Trdata.tmp)
head(Trdata.tmp)

Trdata <- Trdata.tmp %>% 
  select(scientific2, diet.5cat, migrant, bodymass.value)
  
table(Trdata$diet.5cat)
Trdata$diet.5cat <- factor(Trdata$diet.5cat, 
                           levels = c("PlantSeed", "Invertebrate", "Omnivore", "VertFishScav"))

Trdata$migrant <- factor(Trdata$migrant, 
                           levels = c("Resident", "Semi", "Migrant" ))

Trdata$log10.bodymass <- log(Trdata$bodymass.value)

# get the species in rows
birds.table <- t(bird_data)

birds.table <- birds.table %>%
  as.data.frame() %>%
  mutate(species = rownames(.)) 

head(birds.table)

# add traits to species 
## double check species order
cbind.data.frame(Trdata$scientific2, birds.table$species)

birds.full.table <- cbind.data.frame(birds.table, Trdata)
# write.csv(birds.full.table, paste0(Model_wd, "/Birds_selection_abutrait_table.csv"), row.names = FALSE)
```

