---
title: "FilteredData"
author: "Aimara Planillo"
date: "2024-01-23"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Repeat the same PCoA and SEMs to the three groups of birds identified in Planillo et al. 2021. DivDist

The analyses in Planillo et al. 2021 also removed all exotic, aquatic and very rare species, as they only created noise in the data, thus using a total of 67 species. Those species were divided into three groups based on their responses to urbanization:

- Group 1: Urban birds - urban exploiters, related to most urbanized areas
- Group 2: Woodland birds - urban tolerants, related to tree cover
- Group 3: Nature birds - urban avoiders, related to nature like areas and limited anthropogenic disturbance

We remove the apus apus and passer domesticus because the difficulties to assess the small scale home range

```{r prepare workspace}
source("./rscripts/source_packages.R")
```

### load data
```{r}
## env covariates grid
covariates_grid <- read.csv("./data/grids_environmental_variables.csv")
rownames(covariates_grid) <- NULL

## env covariates territory centroid - 100m
covariates_terr <- read.csv("./data/territories_environmental_variables.csv") 
rownames(covariates_terr) <- NULL

## env covariates territory points - 20m
covariates_points <- read.csv("./data/locations_environmental_variables_20m.csv")
rownames(covariates_points) <- NULL


## load list of the birds with the assigned group
bird_groups <- read.csv("./data/species_groups_responses.csv")
bird_groups <- bird_groups %>% 
  filter(!species %in% c("A.apus", "P.domesticus")) %>% 
  rename(short_names = species)

## grid data
bird_data <- read.csv("./data/grid_birds_abundance_corrected.csv") %>%
  rename(transect = X) %>% 
  mutate(transect = tolower(transect)) %>% 
  column_to_rownames("transect")
 
ncol(bird_data) #95 species
nrow(bird_groups) #64 species


bird_matrix <- as.matrix(bird_data)

## order species alphabetically
bird_matrix <- bird_matrix %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("species") %>% 
  arrange(species)


## trait data
birds_traits <- read.csv("data/birds_all_traits.csv") 
head(birds_traits)
birds_traits <- birds_traits %>% 
  arrange(species) %>% 
  filter(!species %in% c("Apus apus", "Passer domesticus")) %>% 
  mutate(migrant_num = as.numeric(migrant_num),
         diet_cat = as.factor(diet_cat))%>%
  dplyr::select(-migrant_cat)

# get species names with underscore
name1 <- stringr::word(birds_traits$species, 1)
name2 <- stringr::word(birds_traits$species, 2)

name_firstletter <- stringr::str_sub(birds_traits$species, 1, 1)

full_names <- paste(name1,name2, sep = "_")
short_names <- paste(name_firstletter, name2, sep = ".")
```

### get tables for the groups
```{r}
## get trait data for the groups
traits_matrix <- birds_traits %>%
  mutate(sp2 = species, 
         species = full_names, 
         short_names = short_names) %>% 
  dplyr::select(-sp_synonym) %>% 
  left_join(bird_groups, by = "short_names") %>% 
  filter(!is.na(group))
# write.csv(traits_matrix, "./output/traits_groups.csv", row.names = FALSE)

nrow(traits_matrix) # 64
## trait data
group_list <- split.data.frame(traits_matrix, f = traits_matrix$group)

## get birds abundance by group
bird_abund <- bird_matrix %>% 
  filter(species %in% traits_matrix$species) %>% 
  mutate(group = traits_matrix$group)

bird_abund_g1 <- bird_abund %>% 
  filter(group == "Group1") %>% 
  dplyr::select(-group) %>% 
  column_to_rownames("species") %>% 
  t()

bird_abund_g2 <- bird_abund %>% 
  filter(group == "Group2") %>% 
  dplyr::select(-group) %>% 
  column_to_rownames("species") %>% 
  t()

bird_abund_g3 <- bird_abund %>% 
  filter(group == "Group3") %>% 
  dplyr::select(-group) %>% 
  column_to_rownames("species") %>% 
  t()

bird_abund_list <- list(bird_abund_g1, bird_abund_g2, bird_abund_g3)
names(bird_abund_list) <- c("Group1", "Group2", "Group3")

```


# 1. Calculate FD indices for the different monitoring grids

```{r}
names(group_list)
names(bird_abund_list)

mean_traits_groups <- data.frame()
FD_groups <- data.frame()

for(i in 1:length(group_list)){
  traits_group <- group_list[[i]] 
  rownames(traits_group) <- NULL
  traits_group <- traits_group %>% 
    dplyr::select(-short_names, -sp2) %>% 
    column_to_rownames("species")
  
  abund_group <- bird_abund_list[[i]]   
  abund_group <- abund_group %>% 
    as.matrix()
  
  FD_group <- FD::dbFD(x = traits_group, a = abund_group, 
                     w.abun = TRUE, corr = 'cailliez', 
                     m = 'max', clust.type = 'ward',
                     print.pco = TRUE, calc.FGR = FALSE, ## since we decided we'll focus on FRic, Fdis 
                     stand.FRic = TRUE, calc.FDiv = TRUE)
  
  ## weighted mean of functional traits
  mean_traits <- FD_group$CWM 
  mean_traits <- mean_traits %>% 
    as.data.frame() %>% 
    rownames_to_column("transect") %>% 
    mutate(group = names(group_list)[i]) 

  mean_traits_groups <- rbind(mean_traits_groups, mean_traits)
  # extract diversity values
  birds_fr <- as.data.frame(FD_group$FRic)
  birds_fe <- as.data.frame(FD_group$FEve)
  birds_fd <- as.data.frame(FD_group$FDiv)

  row.names(birds_fr)
  birds_fr$transect <- rownames(birds_fr)
  birds_fe$transect <- rownames(birds_fe)
  birds_fd$transect <- rownames(birds_fd)

  birds_functional_measures <- left_join(birds_fr, birds_fe,by = "transect")
  birds_functional_measures <- left_join(birds_functional_measures, birds_fd, by = "transect")
  head(birds_functional_measures)
  birds_functional_measures <- birds_functional_measures %>% 
    mutate(group = names(group_list)[i]) %>% 
  dplyr::select(transect, group, everything())
  
  FD_groups <- rbind(FD_groups, birds_functional_measures)
}

mean_traits_groups
FD_groups

# write.csv(FD_groups, "./output/functional_measures_transects_groups.csv", row.names = FALSE)
# write.csv(mean_traits_groups, "./output/functional_measures_transects_cwm_groups.csv", row.names = FALSE)
```


### summaries
```{r}
FD_groups %>% 
  group_by(group) %>% 
  summarise(across(where(is.numeric), list(mean = mean, sd = sd, min = min, max = max)))
```
# A tibble: 3 × 13
  group  `FD_group$FRic_mean` `FD_group$FRic_sd` `FD_group$FRic_min` `FD_group$FRic_max` `FD_group$FEve_mean`
  <chr>                 <dbl>              <dbl>               <dbl>               <dbl>                <dbl>
1 Group1                0.550              0.228             0.0140                1                    0.558
2 Group2                0.447              0.316             0.0320                1                    0.620
3 Group3                0.381              0.270             0.00261               0.893                0.710


# Function to run PCoA and extract info

```{r}
source("./rscripts/source_functions.R")
```


## Group 1

### prepare data
```{r load data}
## bird abund transect per group
bird_abund_g1 <- bird_abund_g1 %>% 
  as.data.frame() %>% 
  rownames_to_column("transect") %>% 
  arrange(transect) %>% 
  column_to_rownames("transect")

## get species in g1
species_g1 <- traits_matrix %>% 
  filter(group == "Group1") %>% 
  dplyr::select(sp2)
nrow(species_g1) # 10 sp

## trait data transect
mean_traits_g1 <- mean_traits_groups %>% 
  filter(group == "Group1") %>% 
  arrange(transect) %>% 
  dplyr::select(-group, -nsites) %>% 
  column_to_rownames("transect")

## env covariates grid
covariates_grid_g1 <- covariates_grid %>% 
  filter(transect %in% rownames(bird_abund_g1)) %>% 
  arrange(transect) %>% 
  dplyr::select(-lat, -long) %>% 
  column_to_rownames("transect") 

## env covariates territory
## filtered to species for the group
covariates_terr_g1 <- covariates_terr %>% 
  filter(transect %in% rownames(bird_abund_g1)) %>% 
  filter(sp_name %in% species_g1$sp2) %>% 
  dplyr::select(-nvisits) %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  column_to_rownames("transect") %>%  
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean$", "", .)))

## env covariates points
## filtered to species for the group
covariates_pts_g1 <- covariates_points %>% 
  filter(transect %in% rownames(bird_abund_g1)) %>% 
  filter(sp_name %in% species_g1$sp2) %>% 
  dplyr::select(-visit) %>% 
  group_by(unique_ter_id, transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  ungroup() %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  column_to_rownames("transect") %>%  
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean_mean$", "", .)))  
  
```


### grid scale
```{r}
pcoa_grid_g1 <- mypcoa_toplot(species_matrix = bird_abund_g1, 
                      env_matrix = covariates_grid_g1, 
                      traits_matrix = mean_traits_g1)

plot_grid_g1 <- ggplotting_pcoa(site_scores = pcoa_grid_g1$site_scores,
                                env_coord_cont = pcoa_grid_g1$env_coord_cont, 
                                traits_coord_cat = pcoa_grid_g1$traits_coord_cat,
                                traits_coord_cont = pcoa_grid_g1$traits_coord_cont,
                                myxlim = c(-1.2, 1.2), myylim = c(-1.2, 1),
                                mytitle = "PCoA Grid Scale - Group 1")
plot_grid_g1
# ggsave(plot = plot_grid_g1,
#        filename = "./plots/groups/PCoA_grid_scale_g1.png",
#        dpi = 900, height = 8, width = 8)
# saveRDS(plot_grid_g1, "./output/PCoAplot_grid_group1.rds")
```

### territory centroid scale
```{r}
pcoa_terr_g1 <- mypcoa_toplot(species_matrix = bird_abund_g1, 
                      env_matrix = covariates_terr_g1, 
                      traits_matrix = mean_traits_g1)

plot_terr_g1 <- ggplotting_pcoa(site_scores = pcoa_terr_g1$site_scores,
                                env_coord_cont = pcoa_terr_g1$env_coord_cont, 
                                traits_coord_cat = pcoa_terr_g1$traits_coord_cat,
                                traits_coord_cont = pcoa_terr_g1$traits_coord_cont,
                                myxlim = c(-1.2, 1.2), myylim = c(-1.2, 1),
                                mytitle = "PCoA Territory Centroid Scale - Group 1")
plot_terr_g1
# ggsave(plot = plot_terr_g1, 
#        filename = "./plots/groups/PCoA_terr_scale_g1.png", 
#        dpi = 900, height = 8, width = 8)
# saveRDS(plot_terr_g1, "./output/PCoAplot_terr_group1.rds")
```

### point location scale
```{r}
pcoa_pts_g1 <- mypcoa_toplot(species_matrix = bird_abund_g1, 
                      env_matrix = covariates_pts_g1, 
                      traits_matrix = mean_traits_g1)

plot_pts_g1 <- ggplotting_pcoa(site_scores = pcoa_pts_g1$site_scores,
                               env_coord_cont = pcoa_pts_g1$env_coord_cont, 
                               traits_coord_cat = pcoa_pts_g1$traits_coord_cat,
                               traits_coord_cont = pcoa_pts_g1$traits_coord_cont,
                                myxlim = c(-1.2, 1.2), myylim = c(-1.2, 1),
                               mytitle = "PCoA Point Scale - Group 1")
plot_pts_g1
# ggsave(plot = plot_pts_g1, 
#        filename = "./plots/groups/PCoA_point_scale_g1.png", 
#        dpi = 900, height = 8, width = 8)
# saveRDS(plot_pts_g1, "./output/PCoAplot_pts_group1.rds")
```


## Group 2

### prepare data
```{r load data}
## bird abund transect per group
bird_abund_g2 <- bird_abund_g2 %>% 
  as.data.frame() %>% 
  rownames_to_column("transect") %>% 
  arrange(transect) %>% 
  column_to_rownames("transect")

## get species in g2
species_g2 <- traits_matrix %>% 
  filter(group == "Group2") %>% 
  dplyr::select(sp2)
nrow(species_g2) # 12 sp

## trait data transect
mean_traits_g2 <- mean_traits_groups %>% 
  filter(group == "Group1") %>% 
  arrange(transect) %>% 
  dplyr::select(-group, -nsites) %>% 
  column_to_rownames("transect")

## env covariates grid
covariates_grid_g2 <- covariates_grid %>% 
  filter(transect %in% rownames(bird_abund_g2)) %>% 
  arrange(transect) %>% 
  dplyr::select(-lat, -long) %>% 
  column_to_rownames("transect") 

## env covariates territory

covariates_terr_g2 <- covariates_terr %>% 
  filter(transect %in% rownames(bird_abund_g2)) %>% 
  filter(sp_name %in% species_g2$sp2) %>% 
  dplyr::select(-nvisits) %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  column_to_rownames("transect") %>%  
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean$", "", .)))

## env covariates points
covariates_pts_g2 <- covariates_points %>% 
  filter(transect %in% rownames(bird_abund_g2)) %>% 
  filter(sp_name %in% species_g2$sp2) %>% 
  dplyr::select(-visit) %>% 
  group_by(unique_ter_id, transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  ungroup() %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  column_to_rownames("transect") %>%  
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean_mean$", "", .)))
```


### grid scale
```{r}
pcoa_grid_g2 <- mypcoa_toplot(species_matrix = bird_abund_g2, 
                      env_matrix = covariates_grid_g2, 
                      traits_matrix = mean_traits_g2)

plot_grid_g2 <- ggplotting_pcoa(site_scores = pcoa_grid_g2$site_scores,
                                env_coord_cont = pcoa_grid_g2$env_coord_cont, 
                                traits_coord_cat = pcoa_grid_g2$traits_coord_cat,
                                traits_coord_cont = pcoa_grid_g2$traits_coord_cont,
                                myxlim = c(-1.2, 1.2), myylim = c(-1.2, 1),
                                mytitle = "PCoA Grid Scale - Group 2")
plot_grid_g2
# ggsave(plot = plot_grid_g2, 
#        filename = "./plots/groups/PCoA_grid_scale_g2.png", 
#        dpi = 900, height = 8, width = 8)
# saveRDS(plot_grid_g2, "./output/PCoAplot_grid_group2.rds")
```

### territory centroid scale
```{r}
pcoa_terr_g2 <- mypcoa_toplot(species_matrix = bird_abund_g2, 
                      env_matrix = covariates_terr_g2, 
                      traits_matrix = mean_traits_g2)

plot_terr_g2 <- ggplotting_pcoa(site_scores = pcoa_terr_g2$site_scores,
                                env_coord_cont = pcoa_terr_g2$env_coord_cont, 
                                traits_coord_cat = pcoa_terr_g2$traits_coord_cat,
                                traits_coord_cont = pcoa_terr_g2$traits_coord_cont,
                                myxlim = c(-1.2, 1.2), myylim = c(-1.2, 1),
                                mytitle = "PCoA Territory Centroid Scale - Group 2")
plot_terr_g2
# ggsave(plot = plot_terr_g2, 
#        filename = "./plots/groups/PCoA_terr_scale_g2.png", 
#        dpi = 900, height = 8, width = 8)
# saveRDS(plot_terr_g2, "./output/PCoAplot_terr_group2.rds")
```

### point location scale
```{r}
pcoa_pts_g2 <- mypcoa_toplot(species_matrix = bird_abund_g2, 
                      env_matrix = covariates_pts_g2, 
                      traits_matrix = mean_traits_g2)

plot_pts_g2 <- ggplotting_pcoa(site_scores = pcoa_pts_g2$site_scores,
                                env_coord_cont = pcoa_pts_g2$env_coord_cont, 
                                traits_coord_cat = pcoa_pts_g2$traits_coord_cat,
                                traits_coord_cont = pcoa_pts_g2$traits_coord_cont,
                               myxlim = c(-1.2, 1.2), myylim = c(-1.2, 1),
                               mytitle = "PCoA Point Scale - Group 2")
plot_pts_g2
# ggsave(plot = plot_pts_g2, 
#        filename = "./plots/groups/PCoA_point_scale_g2.png", 
#        dpi = 900, height = 8, width = 8)
# saveRDS(plot_pts_g2, "./output/PCoAplot_pts_group2.rds")
```


## Group 3

### prepare data
```{r load data}
## bird abund transect per group
bird_abund_g3 <- bird_abund_g3 %>% 
  as.data.frame() %>% 
  rownames_to_column("transect") %>% 
  arrange(transect) %>% 
  column_to_rownames("transect")

## get species in g2
species_g3 <- traits_matrix %>% 
  filter(group == "Group3") %>% 
  dplyr::select(sp2)
nrow(species_g3) # 42 sp

## trait data transect
mean_traits_g3 <- mean_traits_groups %>% 
  filter(group == "Group1") %>% 
  arrange(transect) %>% 
  dplyr::select(-group, -nsites) %>% 
  column_to_rownames("transect")

## env covariates grid
covariates_grid_g3 <- covariates_grid %>% 
  filter(transect %in% rownames(bird_abund_g3)) %>% 
  arrange(transect) %>% 
  dplyr::select(-lat, -long) %>% 
  column_to_rownames("transect") 

## env covariates territory
 # sum(unique(covariates_terr$sp_name) %in% species_g3$sp2)
covariates_terr_g3 <- covariates_terr %>% 
  filter(transect %in% rownames(bird_abund_g3)) %>% 
  filter(sp_name %in% species_g2$sp2) %>% 
  dplyr::select(-nvisits) %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  column_to_rownames("transect") %>%  
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean$", "", .)))

## env covariates points
covariates_pts_g3 <- covariates_points %>% 
  filter(transect %in% rownames(bird_abund_g3)) %>% 
  filter(sp_name %in% species_g2$sp2) %>% 
  dplyr::select(-visit) %>% 
  group_by(unique_ter_id, transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  ungroup() %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  column_to_rownames("transect") %>%  
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean_mean$", "", .)))
```


### grid scale
```{r}
pcoa_grid_g3 <- mypcoa_toplot(species_matrix = bird_abund_g3, 
                      env_matrix = covariates_grid_g3, 
                      traits_matrix = mean_traits_g3)

plot_grid_g3 <- ggplotting_pcoa(site_scores = pcoa_grid_g3$site_scores,
                                env_coord_cont = pcoa_grid_g3$env_coord_cont, 
                                traits_coord_cat = pcoa_grid_g3$traits_coord_cat,
                                traits_coord_cont = pcoa_grid_g3$traits_coord_cont,
                                myxlim = c(-1.2, 1.2), myylim = c(-1.2, 1),
                                mytitle = "PCoA Grid Scale - Group 3")
plot_grid_g3
# ggsave(plot = plot_grid_g3, 
#        filename = "./plots/groups/PCoA_grid_scale_g3.png", 
#        dpi = 900, height = 8, width = 8)
# saveRDS(plot_grid_g3, "./output/PCoAplot_grid_group3.rds")
```

### territory centroid scale
```{r}
pcoa_terr_g3 <- mypcoa_toplot(species_matrix = bird_abund_g3, 
                      env_matrix = covariates_terr_g3, 
                      traits_matrix = mean_traits_g3)

plot_terr_g3 <- ggplotting_pcoa(site_scores = pcoa_terr_g3$site_scores,
                                env_coord_cont = pcoa_terr_g3$env_coord_cont, 
                                traits_coord_cat = pcoa_terr_g3$traits_coord_cat,
                                traits_coord_cont = pcoa_terr_g3$traits_coord_cont,
                                myxlim = c(-1.2, 1.2), myylim = c(-1.2, 1),
                                mytitle = "PCoA Territory Centroid Scale - Group 3")
plot_terr_g3
# ggsave(plot = plot_terr_g3, 
#        filename = "./plots/groups/PCoA_terr_scale_g3.png", 
#        dpi = 900, height = 8, width = 8)
# saveRDS(plot_terr_g3, "./output/PCoAplot_terr_group3.rds")
```

### point location scale
```{r}
pcoa_pts_g3 <- mypcoa_toplot(species_matrix = bird_abund_g3, 
                      env_matrix = covariates_pts_g3, 
                      traits_matrix = mean_traits_g3)

plot_pts_g3 <- ggplotting_pcoa(site_scores = pcoa_pts_g3$site_scores,
                                env_coord_cont = pcoa_pts_g3$env_coord_cont, 
                                traits_coord_cat = pcoa_pts_g3$traits_coord_cat,
                                traits_coord_cont = pcoa_pts_g3$traits_coord_cont,
                                myxlim = c(-1.2, 1.2), myylim = c(-1.2, 1),
                                mytitle = "PCoA Point Scale - Group 3")
plot_pts_g3
# ggsave(plot = plot_pts_g3, 
#        filename = "./plots/groups/PCoA_point_scale_g3.png", 
#        dpi = 900, height = 8, width = 8)
# saveRDS(plot_pts_g3, "./output/PCoAplot_pts_group3.rds")
```



# Put all plots together
Including the ones with all the birds

!!This works only on the same computer that the plots were saved

```{r load data, echo = FALSE}
#library(patchwork)

myplot_grid <- readRDS("./output/PCoAplot_grid_allbirds.rds")
myplot_terr <- readRDS("./output/PCoAplot_terr_allbirds.rds")
myplot_pts <- readRDS("./output/PCoAplot_pts_allbirds.rds")

group1_grid <- readRDS("./output/PCoAplot_grid_group1.rds")
group1_terr <- readRDS("./output/PCoAplot_terr_group1.rds")
group1_pts <- readRDS("./output/PCoAplot_pts_group1.rds")

group2_grid <- readRDS("./output/PCoAplot_grid_group2.rds")
group2_terr <- readRDS("./output/PCoAplot_terr_group2.rds")
group2_pts <- readRDS("./output/PCoAplot_pts_group2.rds")

group3_grid <- readRDS("./output/PCoAplot_grid_group3.rds")
group3_terr <- readRDS("./output/PCoAplot_terr_group3.rds")
group3_pts <- readRDS("./output/PCoAplot_pts_group3.rds")
```

```{r prepare plots}
## get plots without titles
plot_grid <- myplot_grid + ggtitle(NULL) +
  labs(tag = "A") +
  # xlab("MDS1") +
  # ylab("MDS2") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.16, 0.95))

plot_terr <- myplot_terr + ggtitle(NULL) +
  labs(tag = "B") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.12, 0.95))

plot_pts <- myplot_pts + ggtitle(NULL) +
  labs(tag = "C") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.12, 0.95))


plot_g1_grid <- group1_grid + ggtitle(NULL) +
  labs(tag = "D") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.16, 0.95))

plot_g1_terr <- group1_terr + ggtitle(NULL) +
  labs(tag = "E") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.12, 0.95))

plot_g1_pts <- group1_pts + ggtitle(NULL) +
  labs(tag = "F") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.12, 0.95))


plot_g2_grid <- group2_grid + ggtitle(NULL) +
  labs(tag = "G") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.16, 0.95))

plot_g2_terr <- group2_terr + ggtitle(NULL) +
  labs(tag = "H") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.12, 0.95))

plot_g2_pts <- group2_pts + ggtitle(NULL) +
  labs(tag = "I") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.12, 0.95))


plot_g3_grid <- group3_grid + ggtitle(NULL) +
  labs(tag = "J") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.16, 0.95))

plot_g3_terr <- group3_terr + ggtitle(NULL) +
  labs(tag = "K") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.12, 0.95))

plot_g3_pts <- group3_pts + ggtitle(NULL) +
  labs(tag = "L") +
  theme(plot.tag = element_text(hjust = 0, size = 14, vjust = 0, face = "bold"),
    plot.tag.position = c(0.12, 0.95))


## get row names
col1 <- ggplot() + annotate(geom = 'text', x=1, y=1, label="Grid Scale", size = 8) + theme_void() 
col2 <- ggplot() + annotate(geom = 'text', x=1, y=1, label="Territory Scale", size = 8) + theme_void() 
col3 <- ggplot() + annotate(geom = 'text', x=1, y=1, label="Point Scale", size = 8) + theme_void() 
row1 <- ggplot() + annotate(geom = 'text', x=1, y=1, label="All birds", angle = 90, size = 8) + theme_void() 
row2 <- ggplot() + annotate(geom = 'text', x=1, y=1, label="Exploiters", angle = 90, size = 8) + theme_void() 
row3 <- ggplot() + annotate(geom = 'text', x=1, y=1, label="Adapters", angle = 90, size = 8) + theme_void() 
row4 <- ggplot() + annotate(geom = 'text', x=1, y=1, label="Avoiders", angle = 90, size = 8) + theme_void()
```


```{r prepare plot layout}
## define layout 
layoutplot <- "
#aaabbbccc
dhhhiiijjj
dhhhiiijjj
elllmmmooo
elllmmmooo
fpppqqqrrr
fpppqqqrrr
gssstttvvv
gssstttvvv
"

plotlist <- list(a = col1, b = col2, c = col3,
                 d = row1, e = row2, f = row3, g = row4, 
                 h = plot_grid, i = plot_terr, j = plot_pts,
                 l = plot_g1_grid, m = plot_g1_terr, o = plot_g1_pts, 
                 p = plot_g2_grid, q = plot_g2_terr, r = plot_g2_pts,
                 s = plot_g3_grid, t = plot_g3_terr, v = plot_g3_pts)

wrap_plots(plotlist, guides = 'collect', design = layoutplot) +
  plot_layout(axis_titles = "collect", heights = c(1, 0.1 , 3, 0.1, 3, 0.1, 3, 0.1, 3, 0.1)) 
  # plot_annotation(tag_levels = 'A') 
  
# ggsave(filename = "./plots/PCoA_allplots_allgroups.png", 
#         dpi = 600, height = 20, width = 15)
```


