---
title: "SEM_groups"
output: html_document
date: "2024-05-13"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# SEM models for bird groups

```{r}
source("./rscripts/source_packages.R")
```


### load data
```{r}
## FD data
FD_groups <- read.csv("./output/functional_measures_transects_groups.csv")

## env covariates grid
covariates_grid <- read.csv("./data/grids_environmental_variables.csv")

## env covariates territory centroid - 100m
covariates_terr <- read.csv("./data/territories_environmental_variables.csv") 

## env covariates territory points - 20m
covariates_points <- read.csv("./data/locations_environmental_variables_20m.csv")

## load the trait data to get species names
traits_groups <- read.csv("./output/traits_groups.csv")
```

### scale FD values
```{r}
## standardize values
fd_groups_std <- FD_groups %>% 
  group_by(group) %>% 
  arrange(transect) %>% 
  mutate(across(where(is.numeric), list(scale))) %>% 
  ungroup()
  
colnames(fd_groups_std)[6:8] <- c("birds_fric_std", "birds_feve_std", "birds_fdiv_std")
  
fd_g1 <- filter(fd_groups_std, group == "Group1")
fd_g2 <- filter(fd_groups_std, group == "Group2")
fd_g3 <- filter(fd_groups_std, group == "Group3")

summary(fd_g3)
```


### prepare groups and associated env data
```{r get species and transects for each group}
## transects per group
transect_g1 <- FD_groups %>% 
  filter(group == "Group1") %>% 
  dplyr::select(transect)

transect_g2 <- FD_groups %>% 
  filter(group == "Group2") %>% 
  dplyr::select(transect)

transect_g3 <- FD_groups %>% 
  filter(group == "Group3") %>% 
  dplyr::select(transect)

## species in each transect## species in each group
species_g1 <- traits_groups %>% 
  filter(group == "Group1") %>% 
  dplyr::select(sp2)
nrow(species_g1) # 10 sp

species_g2 <- traits_groups %>% 
  filter(group == "Group2") %>% 
  dplyr::select(sp2)
nrow(species_g2) # 12sp

species_g3 <- traits_groups %>% 
  filter(group == "Group3") %>% 
  dplyr::select(sp2)
nrow(species_g3) # 42 sp
```


```{r group data grid scale}
## env covariates grid 
covariates_grid_g1_std <- covariates_grid %>% 
  filter(transect %in% transect_g1$transect) %>% 
  arrange(transect) %>% 
  dplyr::select(-lat, -long) %>% 
  mutate(across(where(is.numeric), list(std = scale)))

covariates_grid_g2_std <- covariates_grid %>% 
  filter(transect %in% transect_g2$transect) %>% 
  arrange(transect) %>% 
  dplyr::select(-lat, -long) %>% 
  mutate(across(where(is.numeric), list(std = scale)))

covariates_grid_g3_std <- covariates_grid %>% 
  filter(transect %in% transect_g3$transect) %>% 
  arrange(transect) %>% 
  dplyr::select(-lat, -long) %>% 
  mutate(across(where(is.numeric), list(std = scale)))
```

```{r group data territory scale}
## env covariates territory
## filtered to species for the group

covariates_terr_g1_std <- covariates_terr %>% 
  filter(transect %in% transect_g1$transect) %>% 
  filter(sp_name %in% species_g1$sp2) %>% 
  dplyr::select(-nvisits) %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean$", "", .))) %>% 
  mutate(across(where(is.numeric), list(std = scale)))

covariates_terr_g2_std <- covariates_terr %>% 
  filter(transect %in% transect_g2$transect) %>% 
  filter(sp_name %in% species_g2$sp2) %>% 
  dplyr::select(-nvisits) %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean$", "", .))) %>% 
  mutate(across(where(is.numeric), list(std = scale)))

covariates_terr_g3_std <- covariates_terr %>% 
  filter(transect %in% transect_g3$transect) %>% 
  filter(sp_name %in% species_g3$sp2) %>% 
  dplyr::select(-nvisits) %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean$", "", .))) %>% 
  mutate(across(where(is.numeric), list(std = scale)))
```

```{r group data point scale}
## env covariates points
## filtered to species for the group
covariates_pts_g1_std <- covariates_points %>% 
  filter(transect %in% transect_g1$transect) %>% 
  filter(sp_name %in% species_g1$sp2) %>% 
  dplyr::select(-visit) %>% 
  group_by(unique_ter_id, transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  ungroup() %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean_mean$", "", .))) %>% 
    mutate(across(where(is.numeric), list(std = scale)))

covariates_pts_g2_std <- covariates_points %>% 
  filter(transect %in% transect_g2$transect) %>% 
  filter(sp_name %in% species_g2$sp2) %>% 
  dplyr::select(-visit) %>% 
  group_by(unique_ter_id, transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  ungroup() %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean_mean$", "", .))) %>% 
    mutate(across(where(is.numeric), list(std = scale)))

covariates_pts_g3_std <- covariates_points %>% 
  filter(transect %in% transect_g3$transect) %>% 
  filter(sp_name %in% species_g3$sp2) %>% 
  dplyr::select(-visit) %>% 
  group_by(unique_ter_id, transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  ungroup() %>% 
  group_by(transect) %>% 
  summarise(across(where(is.numeric), list(mean = ~mean(., na.rm = TRUE)))) %>% 
  arrange(transect) %>% 
  ungroup() %>% 
  rename_at(.vars = vars(ends_with("_mean")),
            .funs = list(~sub("_mean_mean$", "", .))) %>% 
    mutate(across(where(is.numeric), list(std = scale)))

```

### prepare data tables to run the models 
```{r}
## grid data
g1_grid_data <- left_join(covariates_grid_g1_std, fd_g1, by = "transect")
str(g1_grid_data)

g2_grid_data <- left_join(covariates_grid_g2_std, fd_g2, by = "transect")
str(g2_grid_data)

g3_grid_data <- left_join(covariates_grid_g3_std, fd_g3, by = "transect")
str(g3_grid_data)

groups_grid_data <- list(g1_grid_data, g2_grid_data, g3_grid_data)
names(groups_grid_data) <- c("grid_group1", "grid_group2", "grid_group3")

## territory data
g1_terr_data <- left_join(covariates_terr_g1_std, fd_g1, by = "transect")
str(g1_grid_data)
g2_terr_data <- left_join(covariates_terr_g2_std, fd_g2, by = "transect")
g3_terr_data <- left_join(covariates_terr_g3_std, fd_g3, by = "transect")

groups_terr_data <- list(g1_terr_data, g2_terr_data, g3_terr_data)
names(groups_terr_data) <- c("terr_group1", "terr_group2", "terr_group3")

## point data
g1_point_data <- left_join(covariates_pts_g1_std, fd_g1, by = "transect")
str(g1_grid_data)
g2_point_data <- left_join(covariates_pts_g2_std, fd_g2, by = "transect")
g3_point_data <- left_join(covariates_pts_g3_std, fd_g3, by = "transect")

groups_point_data <- list(g1_point_data, g2_point_data, g3_point_data)
names(groups_point_data) <- c("point_group1", "point_group2", "point_group3")

## all data together
groups_all_data <- c(groups_grid_data, groups_terr_data, groups_point_data)
length(groups_all_data)
```


# FUNCTIONAL EVENNESS - FEVE

### define model
```{r}
feve_sem_groups <-sem(
  ' # direct effects
  birds_feve_std ~ veg_pca1_std + veg_pca2_std + glcm_hmg_std + pop_den_std + noise_std

  # mediators
  veg_pca1_std ~ pop_den_std
  veg_pca1_std ~ noise_std
  veg_pca2_std ~ pop_den_std
  veg_pca2_std ~ noise_std
  glcm_hmg_std ~ pop_den_std
  glcm_hmg_std ~ noise_std
  
  # variances of exogenous variables
  veg_pca1_std ~~ veg_pca1_std
  veg_pca2_std ~~ veg_pca2_std
  glcm_hmg_std ~~ glcm_hmg_std
  noise_std    ~~ noise_std
  pop_den_std  ~~ pop_den_std
  
  # covariance of exogenouse variables
  veg_pca1_std ~~ glcm_hmg_std
  veg_pca2_std ~~ glcm_hmg_std
  ')
```

## run feve models for all groups and scales in loop 
```{r}
## fit of the models
fit_feve_list <- list()

for(i in 1:length(groups_all_data)){
  my_scale <- names(groups_all_data)[i]
  my_data <- groups_all_data[[i]]
  
  fit_feve_tmp <- lavaan::sem(feve_sem_groups, my_data)
  
  fit_feve_list[[i]] <- fit_feve_tmp
  names(fit_feve_list)[i] <- my_scale
}

length(fit_feve_list) # 9 (3 groups * 3 scales)
names(fit_feve_list)

summary(fit_feve_list$grid_group1,
        rsquare = T,
        standardized = T,
        fit.measures = T)
# saveRDS(fit_feve_list, "./output/sem_groups/SEM_feve_allgroups.RDS")

## evaluate the models
feve_models_cfi_values_table <- data.frame()
feve_models_parameter_list <- list()
feve_models_parameters_table <- data.frame()

for(i in 1:length(fit_feve_list)){
  my_model <- fit_feve_list[[i]]
  model_name <- names(fit_feve_list)[i]
  ## get cfi value
  cfi_model <- fitmeasures(my_model)["cfi"]
  cfi_tmp <- cbind.data.frame(model_name, cfi_model)
  feve_models_cfi_values_table <- rbind(feve_models_cfi_values_table, cfi_tmp)
  
  ## get model parameters
  table_parameters <- parameterEstimates(my_model, standardized=TRUE) %>%  head(11)
  feve_models_parameter_list[[i]] <- table_parameters
  names(feve_models_parameter_list)[i] <- model_name
  
   # in a table
  table_parameters_df <- table_parameters %>% 
    mutate(scale = model_name)
  
  feve_models_parameters_table <- rbind(feve_models_parameters_table, table_parameters_df)
}

feve_models_cfi_values_table
 # write.csv(feve_models_cfi_values_table, "./output/sem_groups/SEM_feve_allgroups_cfi_values.csv", row.names = FALSE)
feve_models_parameter_list
 # saveRDS(feve_models_parameter_list, "./output/sem_groups/SEM_feve_allgroups_parameterTable.rds")
feve_models_parameters_table
# write.csv(feve_models_parameters_table, "./output/sem_groups/SEM_feve_allgroups_parameterTable_df.csv", row.names = FALSE)
```


# FUNCTIONAL RICHNESS - FRIC

### define model
```{r}
fric_sem_groups <-sem(
  ' # direct effects
  birds_fric_std ~ veg_pca1_std + veg_pca2_std + glcm_hmg_std + pop_den_std + noise_std
  
  # mediators
  veg_pca1_std ~ pop_den_std
  veg_pca1_std ~ noise_std
  veg_pca2_std ~ pop_den_std
  veg_pca2_std ~ noise_std
  glcm_hmg_std ~ pop_den_std
  glcm_hmg_std ~ noise_std
  
  # variances of exogenous variables
  veg_pca1_std ~~ veg_pca1_std
  veg_pca2_std ~~ veg_pca2_std
  glcm_hmg_std ~~ glcm_hmg_std
  noise_std    ~~ noise_std
  pop_den_std  ~~ pop_den_std
  
  # covariance of exogenouse variables
  veg_pca1_std ~~ glcm_hmg_std
  veg_pca2_std ~~ glcm_hmg_std
  ')
```

## run fric models for all groups and scales in loop 
```{r}
## fit of the models
fit_fric_list <- list()

for(i in 1:length(groups_all_data)){
  my_scale <- names(groups_all_data)[i]
  my_data <- groups_all_data[[i]]
  
  fit_fric_tmp <- lavaan::sem(fric_sem_groups, my_data)
  
  fit_fric_list[[i]] <- fit_fric_tmp
  names(fit_fric_list)[i] <- my_scale
}

length(fit_fric_list) # 9 (3 groups * 3 scales)
names(fit_fric_list)

summary(fit_fric_list$grid_group1,
        rsquare = T,
        standardized = T,
        fit.measures = T)
# saveRDS(fit_fric_list, "./output/sem_groups/SEM_fric_allgroups.RDS")

## evaluate the models
fric_models_cfi_values_table <- data.frame()
fric_models_parameter_list <- list()
fric_models_parameters_table <- data.frame()

for(i in 1:length(fit_fric_list)){
  my_model <- fit_fric_list[[i]]
  model_name <- names(fit_fric_list)[i]
  ## get cfi value
  cfi_model <- fitmeasures(my_model)["cfi"]
  cfi_tmp <- cbind.data.frame(model_name, cfi_model)
  fric_models_cfi_values_table <- rbind(fric_models_cfi_values_table, cfi_tmp)
  
  ## get model parameters
  table_parameters <- parameterEstimates(my_model, standardized=TRUE) %>%  head(11)
  fric_models_parameter_list[[i]] <- table_parameters
  names(fric_models_parameter_list)[i] <- model_name
  
  # in a table
  table_parameters_df <- table_parameters %>% 
    mutate(scale = model_name)
  
  fric_models_parameters_table <- rbind(fric_models_parameters_table, table_parameters_df)
}

fric_models_cfi_values_table
# write.csv(fric_models_cfi_values_table, "./output/sem_groups/SEM_fric_allgroups_cfi_values.csv", row.names = FALSE)
fric_models_parameter_list
# saveRDS(fric_models_parameter_list, "./output/sem_groups/SEM_fric_allgroups_parameterTable.rds")
fric_models_parameters_table
# write.csv(fric_models_parameters_table, "./output/sem_groups/SEM_fric_allgroups_parameterTable_df.csv", row.names = FALSE)
```


# FUNCTIONAL DIVERGENCE - FDIV

### define model
```{r}
fdiv_sem_groups <-sem(
  ' # direct effects
  birds_fdiv_std ~ veg_pca1_std + veg_pca2_std + glcm_hmg_std + pop_den_std + noise_std
  
  # mediators
  veg_pca1_std ~ pop_den_std
  veg_pca1_std ~ noise_std
  veg_pca2_std ~ pop_den_std
  veg_pca2_std ~ noise_std
  glcm_hmg_std ~ pop_den_std
  glcm_hmg_std ~ noise_std
  
  # variances of exogenous variables
  veg_pca1_std ~~ veg_pca1_std
  veg_pca2_std ~~ veg_pca2_std
  glcm_hmg_std ~~ glcm_hmg_std
  noise_std    ~~ noise_std
  pop_den_std  ~~ pop_den_std
  
  # covariance of exogenouse variables
  veg_pca1_std ~~ glcm_hmg_std
  veg_pca2_std ~~ glcm_hmg_std
  ')
```

## run fdiv models for all groups and scales in loop 
```{r}
## fit of the models
fit_fdiv_list <- list()

for(i in 1:length(groups_all_data)){
  my_scale <- names(groups_all_data)[i]
  my_data <- groups_all_data[[i]]
  
  fit_fdiv_tmp <- lavaan::sem(fdiv_sem_groups, my_data)
  
  fit_fdiv_list[[i]] <- fit_fdiv_tmp
  names(fit_fdiv_list)[i] <- my_scale
}

length(fit_fdiv_list) # 9 (3 groups * 3 scales)
names(fit_fdiv_list)

summary(fit_fdiv_list$grid_group1,
        rsquare = T,
        standardized = T,
        fit.measures = T)
# saveRDS(fit_fdiv_list, "./output/sem_groups/SEM_fdiv_allgroups.RDS")

## evaluate the models
fdiv_models_cfi_values_table <- data.frame()
fdiv_models_parameter_list <- list()
fdiv_models_parameters_table <- data.frame()

for(i in 1:length(fit_fdiv_list)){
  my_model <- fit_fdiv_list[[i]]
  model_name <- names(fit_fdiv_list)[i]
  ## get cfi value
  cfi_model <- fitmeasures(my_model)["cfi"]
  cfi_tmp <- cbind.data.frame(model_name, cfi_model)
  fdiv_models_cfi_values_table <- rbind(fdiv_models_cfi_values_table, cfi_tmp)
  
  ## get model parameters in list
  table_parameters <- parameterEstimates(my_model, standardized=TRUE) %>%  head(11)
  fdiv_models_parameter_list[[i]] <- table_parameters
  names(fdiv_models_parameter_list)[i] <- model_name
  
  # in a table
  table_parameters_df <- table_parameters %>% 
    mutate(scale = model_name)
  
  fdiv_models_parameters_table <- rbind(fdiv_models_parameters_table, table_parameters_df)
}

fdiv_models_cfi_values_table
# write.csv(fdiv_models_cfi_values_table, "./output/sem_groups/SEM_fdiv_allgroups_cfi_values.csv", row.names = FALSE)
fdiv_models_parameter_list
# saveRDS(fdiv_models_parameter_list, "./output/sem_groups/SEM_fdiv_allgroups_parameterTable.rds")
fdiv_models_parameters_table
# write.csv(fdiv_models_parameters_table, "./output/sem_groups/SEM_fdiv_allgroups_parameterTable_df.csv", row.names = FALSE)
```
